Follow-up Samples

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
update methods section

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cacner, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).  Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) (**insert citation**). 16S rRNA gene sequencing was completed as previously described by Kozich, et al. (**insert citation**).  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Which of each group was randomly assigned to avoid biases based on sample collection location.

@@

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  This was repeated 100 times with the best model after all the repeats chosen to be used on the `r length(rownames(good_metaf))` samples with follow up.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model.  The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.

***Selection of Important OTUs***
In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important.

A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.  This new model OTUs were then compared to the overall lesion model and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.

>>>>

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).  Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) (**insert citation**). 16S rRNA gene sequencing was completed as previously described by Kozich, et al. (**insert citation**).  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

@@

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimiation of the mtry hyperparameter as well as data on the best, middle, and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run through 20 repeated 10-fold cross validations to optimzie the model's AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which testing on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.

***Selection of Important OTUs***
In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion prediction model.

A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.  These OTUs were considered the most important for predicting initial and follow up samples.  This new model important OTUs were then compared to the overall lesion model important OTUs and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.

### 2016-12-07 submission/manuscript_outline_20161024.tex
update methods section

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
(\textbf{insert citation}). In brief, study exclusion involved those who
had already undergone surgery, radiation, or chemotherapy, had
colorectal cancer before a baseline stool sample could be obtained, had
IBD, a known hereditary non-polyposis colorectal cacner, or Familial
adenomatous polyposis. Samples used to build the model used for
prediction were collected either prior to a colonoscopy or between 1 - 2
weeks after. The bacterial microbiome has been shown to nomralize within
this time period (\textbf{insert citation}). Kept apart from this
training set were a total of 67 individuals that not only had a sample
as described previoulsy but also a follow up sample between 188 - 546
days after surgery and treatment had been completed. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
citation}). 16S rRNA gene sequencing was completed as previously
described by Kozich, et al. (\textbf{insert citation}). In brief, DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with control, adenoma, and carcinoma
evenly represented on each run. Which of each group was randomly
assigned to avoid biases based on sample collection location.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the 67 individuals with follow up
samples. The model included data on FIT, the bacterial microbiome, sex,
age, Body Mass Index (BMI), whether the indivdiual was caucasion or not,
history of cancer, and family history of cancer. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimiation of the mtry hyperparameter as well
as data on the best and worse performance of the model involved taking
the 490 samples and making 100 80/20 (train/test) splits in the data
where control and lesion were equally represented in the 80 and 20
split, respectively. This 80\% portion was then split again into an
80/20 split, and run twenty times through 10-fold cross validation to
optimzie the model's Receiver Operator Characteristic (ROC) and tuneing
both mtry and ntree. This resulting model was then tested on the 20\% of
the data that was originally held out from this overall process. This
was repeated 100 times with the best model after all the repeats chosen
to be used on the 67 samples with follow up. Once the ideal mtry was
found the entire 490 sample set was used to create the final Random
Forest model. The default cutoff of a probability of 0.5 was used as the
threshold to classify individuals as positive or negative of lesion.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important.

A second model using the same types of conditions was used to identify
OTUs that could predict initial and follow up samples. This new model
OTUs were then compared to the overall lesion model and the resulting
common OTUs to both models were selected to test for changes between
initial and follow up using a paired wilcoxson rank sum test.

>>>>

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
(\textbf{insert citation}). In brief, study exclusion involved those who
had already undergone surgery, radiation, or chemotherapy, had
colorectal cancer before a baseline stool sample could be obtained, had
IBD, a known hereditary non-polyposis colorectal cancer, or Familial
adenomatous polyposis. Samples used to build the model used for
prediction were collected either prior to a colonoscopy or between 1 - 2
weeks after. The bacterial microbiome has been shown to nomralize within
this time period (\textbf{insert citation}). Kept apart from this
training set were a total of 67 individuals that not only had a sample
as described previoulsy but also a follow up sample between 188 - 546
days after surgery and treatment had been completed. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
citation}). 16S rRNA gene sequencing was completed as previously
described by Kozich, et al. (\textbf{insert citation}). In brief, DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with control, adenoma, and carcinoma
evenly represented on each run. Each group was randomly assigned to
avoid biases based on sample collection location.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the 67 individuals with follow up
samples. The model included data on FIT, the bacterial microbiome, sex,
age, Body Mass Index (BMI), whether the indivdiual was caucasion or not,
history of cancer, and family history of cancer. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimiation of the mtry hyperparameter as well
as data on the best, middle, and worse performance of the model involved
taking the 490 samples and making 100 80/20 (train/test) splits in the
data where control and lesion were equally represented in the 80 and 20
split, respectively. This 80\% portion was then split again into an
80/20 split, and run through 20 repeated 10-fold cross validations to
optimzie the model's AUC (Area Under the Curve of the Receiver Operator
Characteristic). This resulting model was then tested on the 20\% of the
data that was originally held out from this overall process. Once the
ideal mtry was found the entire 490 sample set was used to create the
final Random Forest model on which testing on the 67-person cohort was
completed. The default cutoff of 0.5 was used as the threshold to
classify individuals as positive or negative for lesion.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important for the lesion prediction model.

A second model using the same types of conditions was used to identify
OTUs that could predict initial and follow up samples. These OTUs were
considered the most important for predicting initial and follow up
samples. This new model important OTUs were then compared to the overall
lesion model important OTUs and the resulting common OTUs to both models
were selected to test for changes between initial and follow up using a
paired wilcoxson rank sum test.
